Thermodynamic response functions in a cell fluid model

### 2024-08-27 main_morse_response_functions.tex
add reference to Stroker and Meier - Classical stat mech in grand canonical ensemble
rewrite definition of thermal expansion coeff to be suitable for mu and T variables; introduce reduced thermal expansion coefficient

	\subsection{Thermal pressure coefficient}
	The thermal pressure coefficient is defined by
	\begin{equation}
		\label{def:therm_pres_coef}
		\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,\langle N \rangle}.
	\end{equation}
	We rewrite $\beta_V$ into a form that is suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray}
		\beta_V & = & \left(\frac{\partial P}{\partial T}\right)_{V, N}
		\nonumber\\
		& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
	\end{eqnarray}
	Applying the chain rule
	\begin{equation*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
		\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
		= -1
	\end{equation*}
	we obtain
	\begin{eqnarray*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		& = & 
		- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
		\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
		& = & - \left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}.		
	\end{eqnarray*}
	Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
	\begin{equation}
		\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
		- \left(\frac{\partial P}{\partial \mu}\right)_T 
		\left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}
	\end{equation}
	which is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.

	We introduce the reduced thermal pressure coefficient by
	\begin{equation}
		\beta^*_V = \frac{v\beta_V}{k_{\rm B}}.
	\end{equation}

@@

	\subsection{Thermal expansion coefficient}
	The thermal expansion coefficient is defined by
	\begin{equation}
		\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
	\end{equation}

	\begin{figure}[h!]
		\centering \includegraphics[width=0.5\textwidth]{f4a.pdf}
		\vskip-3mm\caption{The thermal expansion coefficient $\alpha_P$ as a function of the density $\bar n$ at different values of relative temperature $\tau > 0$ ($T > T_c$). 
		}\label{fig4a}
	\end{figure}

		\begin{figure}[h!]
		\centering \includegraphics[width=0.45\textwidth]{f4b.pdf}
		\includegraphics[width=0.475\textwidth]{f4c.pdf}
		\vskip-3mm\caption{The thermal expansion coefficient $\alpha_P$ as a function of the effective chemical potential $M$ for different temperatures $\tau = (T - T_c)/T_c$ at $T > T_c$. The two figures differ in the scale of $M$. The left-hand-side figure covers a wider range of $M$, while the right-hand-side one focuses on a range of $M$ around its critical value $0$.
		}\label{fig4b}
	\end{figure}

>>>>

	\subsection{Thermal pressure coefficient}
	The thermal pressure coefficient is defined by
	\begin{equation}
		\label{def:therm_pres_coef}
		\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,\langle N \rangle}.
	\end{equation}
	We rewrite $\beta_V$ into a form that is suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray}
		\beta_V & = & \left(\frac{\partial P}{\partial T}\right)_{V, N}
		\nonumber\\
		& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
	\end{eqnarray}
	Applying the chain rule
	\begin{equation*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
		\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
		= -1
	\end{equation*}
	we obtain
	\begin{eqnarray*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		& = & 
		- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
		\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
		& = & - \left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}.		
	\end{eqnarray*}
	Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
	\begin{equation}
		\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
		- \left(\frac{\partial P}{\partial \mu}\right)_T 
		\left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}
	\end{equation}
	which is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
	(NB. Compare this equation with Eq.(17) from~\cite{StrokerMeier2021})

	We introduce the reduced thermal pressure coefficient by
	\begin{equation}
		\beta^*_V = \frac{v}{k_{\rm B}}\beta_V.
	\end{equation}

@@

	\subsection{Thermal expansion coefficient}
	The thermal expansion coefficient is defined by
	\begin{equation}
		\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
	\end{equation}
	We rewrite $\alpha_P$ into a form suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray*}
		\alpha_P & = & \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
		\\
		& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{P, N}
		\\
		& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{\mu} 
		 - \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
		 \left(\frac{\partial \mu}{\partial T}\right)_{P, N}.
	\end{eqnarray*}
	Applying the chain rule
	\begin{equation*}
		\left(\frac{\partial \mu}{\partial T}\right)_{P, N}
		\left(\frac{\partial T}{\partial P}\right)_{\mu, N}
		\left(\frac{\partial P}{\partial \mu}\right)_{T, N}
		= -1
	\end{equation*}
	we get
	\begin{eqnarray*}
		\left(\frac{\partial \mu}{\partial T}\right)_{P, N} & = & 
		- \left(\frac{\partial P}{\partial T}\right)_{\mu}
		\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}.
	\end{eqnarray*}
	Substituting the result into the expression for $\alpha_P$, we arrive at the final formula for the thermal expansion coefficient
	\begin{equation}
		\alpha_P = -\frac{1}{\rho}\left(\frac{\partial \rho}{\partial T}\right)_{\mu}
		+ \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
		\left(\frac{\partial P}{\partial T}\right)_{\mu}
		\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}
	\end{equation}
	which is readily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
	(NB. Compare this equation with Eqs.(57)-(58) from~\cite{StrokerMeier2021})

	We introduce the reduced thermal expansion coefficient by
	\begin{equation}
		\alpha^*_P = \frac{\varepsilon}{k_{\rm B}} \alpha_P.
	\end{equation}
	Figure~\ref{fig4a} shows the dependency of the reduced thermal expansion coefficient $\alpha^*_P$ on the density $\bar{n}$ for various values of temperature above the critical one. The dependency of $\alpha^*_P$ on the effective chemical potential $M$ is displayed in Figure~\ref{fig4b}.

	\begin{figure}[h!]
		\centering \includegraphics[width=0.5\textwidth]{f4a.pdf}
		\vskip-3mm\caption{The reduced thermal expansion coefficient $\alpha^*_P$ as a function of the density $\bar n$ at different values of relative temperature $\tau > 0$ ($T > T_c$). 
		}\label{fig4a}
	\end{figure}

	\begin{figure}[h!]
		\centering \includegraphics[width=0.45\textwidth]{f4b.pdf}
		\includegraphics[width=0.475\textwidth]{f4c.pdf}
		\vskip-3mm\caption{The reduced thermal expansion coefficient $\alpha^*_P$ as a function of the effective chemical potential $M$ for different temperatures $\tau = (T - T_c)/T_c$ at $T > T_c$. The two figures differ in the scale of $M$. The left-hand-side figure covers a wider range of $M$, while the right-hand-side one focuses on a range of $M$ around its critical value $0$.
		}\label{fig4b}
	\end{figure}
